The genomic impacts of drift and selection for hybrid performance in maize

### 2015-06-11 discussion.tex
intro edits done
jri DONE

In the Iowa RRS experiment, the BSSS and BSCB1 populations steadily lost genetic diversity as they became more differentiated from one another, and principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
This drift was driven by inbreeding through self-pollination and the small effective population sizes used to select a limited number of high-performing, potentially related individuals at each cycle.

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement employed broadly in North America \citep{duvick2004long}.
The trends observed in the Iowa RRS are also a common theme in similar recurrent selection experiments \citep{Romay2013,lamkey2014relative}. 
Increased population structure due to increased inbreeding and lowered effective population size are visible in the modern ‘stiff-stalk’ and ‘non stiff-stalk’ heterotic groups as well \citep{van2012historical}. 
Given the lack of strong selection signals in each case, genetic drift has most likely played a large role in the current genetic structure of modern maize.

Several key inbreds in the ‘stiff-stalk’ heterotic group, B73, B37 and B14, were derived from the BSSS population \citep{darrah19861985, 1986; troyer1999background}. 
B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tabfix} \jri{correct?}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1. 
Both our study and previous work \citep{labate1997molecular} identified genetic differentiation between cycle 0 of BSSS and the founder inbreds. 
The loss of diversity in the cycle 0 population likely led to the loss of some rare alleles and haplotypes. 
The BSSS was maintained for several years before RRS was initiated, and the differentiation could have occurred due to low effective population size during maintenance. 
If so, then this drift will have impacted the trajectory of the RRS program and of the key inbreds derived from BSSS. 
However, cycle 0 has also undergone maintenance since the beginning of the RRS experiment, so some genetic drift may have occurred during this time as well.

Although drift can explain most of the genetic structure genome-wide, phenotypic data provide clear evidence that selection has altered the frequencies of favorable alleles in the BSSS and BSCB1 populations. 
Numerous experiments have shown that the selected populations and the hybrids formed from them are superior to their cycle 0 ancestors \citep{smith1983evaluation, keeratinijakal1993responses, schnicker1993interpopulation, holthaus1995population}. 
Improvement occurred not only for hybrid yield, but concurrently for plant architecture and tolerance to high-density planting in both hybrids and inbreds \citep{brekke2011selection, brekke2011selectionb, edwards2011changes, lauer2012morphological}. 
Since these same trends are observed across wider North American germplasm \citep{tollenaar1989genetic}, identification and characterization of the loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
We find that the genomic regions of extremely low diversity evident at cycle 16 are unlikely to be produced by simulation, and heterozygosity falls more than expected across the genome as a whole.  
However, simulated and observed values are often close, so overall the signature of selection is difficult to detect at any given locus. 
This is due in large part to a limitation of statistical power imposed by inbreeding and low effective population sizes, though uncertainty about the fine-scale recombination map also plays a role. 
We show that an identity-by-descent, haplotype based approach provides additional power as it can distinguish between the fixation of rare and common haplotypes. 
However, the results of single locus simulations are sensitive to sampling error and drift caused by population maintenance. 
In this case, it is difficult to assess significance with only a single population to test. This type of analysis is more effective across several replicated populations, which can control for genetic drift due to the independence between the selections in each replicate \citep{lamkey2014relative}.

\citet{van2012historical} assessed variation in and tested for selection across a wider range of North American maize germplasm using the same SNP array. 
Although selected loci were identified in that study, the overall effects on genomic diversity and haplotype patterns were relatively small compared with the impact of domestication. 
That study was a wide survey across many germplasm sources. 
The limited impact of selection could arise because different haplotypes and loci are selected in different breeding programs. 
In support of this idea, we find that the most likely targets of selection between the BSSS and BSCB1 populations occur at non-overlapping loci. 
This non-overlap between putatively selected loci in complementary populations has also been observed in commercial breeding programs \citep{feng2006temporal}.

>>>>

Our analysis of diversity in the Iowa RRS experiment reveals a steady loss of diversity in the BSSS and BSCB1 populations as they became increasingly differentiated from one another over time.
Principal component analysis shows that as the effective population size and the rates of inbreeding were altered, the rates of change in population structure were altered as well. 
These patterns of population structure, diversity, and differentiation between BSSS and BSCB1 can be largely reproduced by simulation without any selection, supporting the hypothesis that the majority of the genetic structure observed can be attributed to genetic drift alone, despite effective selection for phenotypic improvement. 
This drift was driven by inbreeding through self-pollination and the small effective population sizes used to select a limited number of high-performing, potentially related individuals at each cycle.

Reciprocal recurrent selection serves as a model for the method of hybrid maize improvement employed broadly in North America \citep{duvick2004long}.
The trends observed in the Iowa RRS are also a common theme in similar recurrent selection experiments \citep{romay2012effect,lamkey2014relative}.
Increased population structure due to increased inbreeding and lowered effective population size are visible in the modern ‘stiff-stalk’ and ‘non stiff-stalk’ heterotic groups as well \citep{van2012historical}. 
Given the lack of strong selection signals in each case, genetic drift has most likely played a large role in the current genetic structure of modern maize.

Several key inbreds in the ‘stiff-stalk’ heterotic group, B73, B37 and B14, were derived from the BSSS population \citep{darrah19861985, 1986; troyer1999background}. 
B37 and B14 were derived from cycle 0, and B73 was derived from a half-sib recurrent selection program also started with the BSSS population. 
We examined these three inbreds at the pericentromeric regions listed in Table \ref{tabfix} \jri{correct?}, and found that in most cases they carry different haplotypes from those that rose to high frequency in the RRS experiment.  
Near isogenic lines that substitute these centromeric haplotypes from key inbreds could be used to determine if the haplotypes at high frequency in cycle 16 of BSSS provide a phenotypic advantage when crossed with BSCB1.

Both our study and previous work \citep{labate1997molecular} identified genetic differentiation between cycle 0 of BSSS and the founder inbreds. 
The loss of diversity in the cycle 0 population likely led to the loss of some rare alleles and haplotypes. 
The BSSS was maintained for several years before RRS was initiated, and the differentiation could have occurred due to low effective population size during maintenance. 
If so, then this drift will have impacted the trajectory of the RRS program and of the key inbreds derived from BSSS. 
However, cycle 0 has also undergone maintenance since the beginning of the RRS experiment, so some genetic drift may have occurred during this time as well.

Although drift can explain most of the genetic structure genome-wide, phenotypic data provide clear evidence that selection has altered the frequencies of favorable alleles in the BSSS and BSCB1 populations. 
Numerous experiments have shown that the selected populations and the hybrids formed from them are superior to their cycle 0 ancestors \citep{smith1983evaluation, keeratinijakal1993responses, schnicker1993interpopulation, holthaus1995population}. 
Improvement occurred not only for hybrid yield, but concurrently for plant architecture and tolerance to high-density planting in both hybrids and inbreds \citep{brekke2011selection, brekke2011selectionb, edwards2011changes, lauer2012morphological}. 
Since these same trends are observed across wider North American germplasm \citep{tollenaar1989genetic}, identification and characterization of the loci conferring phenotypic improvement in BSSS and BSCB1 could reveal some of the genetic mechanisms by which maize yield has steadily improved over the decades. 
We find that the genomic regions of extremely low diversity evident at cycle 16 are unlikely to be produced by simulation, and heterozygosity falls more than expected across the genome as a whole.  
However, simulated and observed values are often close, so overall the signature of selection is difficult to detect at any given locus. 
This is due in large part to a limitation of statistical power imposed by inbreeding and low effective population sizes.
\citet{lorenz2015selection} analyze a similar selection experiment with comparable effective population size and also find little little statistical evidence for selection on individual loci in spite of genetic gain for the trait of interest.
Nonetheless, we show that an identity-by-descent, haplotype based approach provides additional power, as it can distinguish between the fixation of rare and common haplotypes. 
However, the results of single locus simulations are sensitive to sampling error and drift caused by population maintenance. 
In this case, it is difficult to assess significance with only a single population to test. This type of analysis is more effective across several replicated populations, which can control for genetic drift due to the independence between the selections in each replicate \citep{lamkey2014relative}.

\citet{van2012historical} assessed variation in and tested for selection across a wider range of North American maize germplasm using the same SNP array. 
Although selected loci were identified in that study, the overall effects on genomic diversity and haplotype patterns were relatively small compared with the impact of domestication. 
That study was a wide survey across many germplasm sources. 
The limited overall impact of selection could arise because different haplotypes and loci are selected in different breeding programs. 
In support of this idea, we find that the most likely targets of selection between the BSSS and BSCB1 populations occur at non-overlapping loci. 
This non-overlap between putatively selected loci in complementary populations has also been observed in commercial breeding programs \citep{feng2006temporal}.
